---
title: "Download climatic data from AEMET"
author: "Fernando"
date: "4/23/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,collapse = TRUE,message = FALSE)
```

```{r libraries}
library(meteoland)
#devtools::install_github("SevillaR/aemet")
library(aemet)
library(lubridate)
library(nngeo)
library(sf)
library(kableExtra)
library(tidyverse)
```

## First step: Get API

Get the API key from AEMET open data service (is free). https://opendata.aemet.es/centrodedescargas/inicio

## API key for AEMET
```{r key}
apikey <- "eyJhbGciOiJIUzI1NiJ9.eyJzdWIiOiJmOGw1aDlAZ21haWwuY29tIiwianRpIjoiNzQ4NzU1ZTgtZDBjYi00MzE3LThmMTUtNzcwZDE0ZGNiNDg0IiwiaXNzIjoiQUVNRVQiLCJpYXQiOjE1ODYxODI2MzIsInVzZXJJZCI6Ijc0ODc1NWU4LWQwY2ItNDMxNy04ZjE1LTc3MGQxNGRjYjQ4NCIsInJvbGUiOiIifQ.Ez8SgcBusQ-514k9vDtEXL29wbRi-cGIYkxnXl5AVv0"
```

## List of dates
Select the set of dates 
```{r}
mydates <- seq(as.Date("2020/3/01"), as.Date("2020/4/20"), "day")
```

## Donwload Meteo Stations

En  la AEMET se puede descargar shp de las estaciones meteorológicas. Están clasificadas en 4 tipos (automáticas, completas, termoétricas, plubiometricas). Aún no se la diferencia entre ellas.

Hasta donde yo se ninguna ofrece datos de humedad.

He utilizado el paquete **aemet** para descargarme información de las estaciones meteorológicas

Para el papel seleccione aquella que me pareció mas correcta y que tuviera datos (hay algunas con muchos datos perdidos)

```{r}
stations <- aemet::aemet_stations(apikey)
stations.sf <- st_as_sf(stations, coords = c("longitud", "latitud"), crs = 4326)
stations.sf$Lon <- stations$longitud
stations.sf$Lat <- stations$latitud
```

## Geometry of provinces

```{r}
load("provinces_spain.RData")
```


## Geometry of Meteo Stations

```{r plot-stations}
ggplot(data = stations.sf) +
  geom_sf(data = provinces_spain, aes()) +
  geom_sf(data = stations.sf, aes(color = provincia), show.legend = FALSE) +
  theme_bw(base_size=6)
```

## Domwload climatic data from each Meteo Station

En el shp "provinces_spain".shp se incluye el campo "Meteo_Station" especificando el ID de la estación seleccionada en cada provincia.

Hay que uncluir un lag temporal. Hay limitación de descargas por minuto.

```{r}
Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[1:40])
```

Extract meteorological data. First initialize the dataframe:
```{r}
df <- data.frame(indicativo = character(), 
                 long = numeric(), 
                 lat = numeric(), 
                 date = ymd(), 
                 Precipitation = numeric(), 
                 MeanTemperature = numeric(), 
                 MinTemperature = numeric(), 
                 MaxTemperature = numeric(), 
                 WindDirection = numeric(), 
                 WindSpeed = numeric(), 
                 SunshineHours = numeric()) 
```

Extract the data tables for these stations and bind to the dataframe:
```{r}
coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
                                                     long = rep(coords$long[1], 51),
                                                     lat = rep(coords$lat[1], each = 51),
                                                     date = Tem1@dates,
                                                     Tem@data[[i]]))}
```

Next set of stations:
```{r}
Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[41:80])
```

Extract the data tables for these stations and bind to the dataframe:
```{r}
coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
                                                     long = rep(coords$long[1], 51),
                                                     lat = rep(coords$lat[1], each = 51),
                                                     date = Tem1@dates,
                                                     Tem@data[[i]]))}
```

Next set of stations:
```{r}
Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[81:120])
```

Extract the data tables for these stations and bind to the dataframe:
```{r}
coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
                                                     long = rep(coords$long[1], 51),
                                                     lat = rep(coords$lat[1], each = 51),
                                                     date = Tem1@dates,
                                                     Tem@data[[i]]))}
```

Next set of stations:
```{r}
Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[121:160])
```

Extract the data tables for these stations and bind to the dataframe:
```{r}
coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
                                                     long = rep(coords$long[1], 51),
                                                     lat = rep(coords$lat[1], each = 51),
                                                     date = Tem1@dates,
                                                     Tem@data[[i]]))}
```

Retrieve the next set of stations:
```{r}
Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[161:200])
```

Extract the data tables for these stations and bind to the dataframe:
```{r}
coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
                                                     long = rep(coords$long[1], 51),
                                                     lat = rep(coords$lat[1], each = 51),
                                                     date = Tem1@dates,
                                                     Tem@data[[i]]))}
```

Retrieve the next set of stations:
```{r}
Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[201:240])
```

Extract the data tables for these stations and bind to the dataframe:
```{r}
coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
                                                     long = rep(coords$long[1], 51),
                                                     lat = rep(coords$lat[1], each = 51),
                                                     date = Tem1@dates,
                                                     Tem@data[[i]]))}
```

Retrieve the next set of stations:
```{r}
Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[241:280])
```

Extract the data tables for these stations and bind to the dataframe:
```{r}
coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
                                                     long = rep(coords$long[1], 51),
                                                     lat = rep(coords$lat[1], each = 51),
                                                     date = Tem1@dates,
                                                     Tem@data[[i]]))}
```

Retrieve the next set of stations:
```{r}
Tem <- downloadAEMEThistorical(apikey, mydates, station_id=stations.sf$indicativo[281:291])
```

Extract the data tables for these stations and bind to the dataframe:
```{r}
coords <- coordinates(Tem) %>% data.frame() %>% rownames_to_column("indicativo")
for(i in 1:nrow(Tem@coords)){df <- rbind(df, 
                                          data.frame(indicativo = rep(coords$indicativo[i], 51), 
                                                     long = rep(coords$long[1], 51),
                                                     lat = rep(coords$lat[1], each = 51),
                                                     date = Tem1@dates,
                                                     Tem@data[[i]]))}
```

Rename dataframe:
```{r}
meteo_data <- df %>% left_join(stations.sf %>% st_drop_geometry() %>% transmute(indicativo, Lon, Lat)) %>%
  select(-c(long, lat))
```

Save to RData file:
```{r}
save(meteo_data, file = "meteo_data.RData")
```


